---
title: "Week 2: Getting started with R"
description: "Using `dplyr` and census data."
date: 2026-09-04
---

## Weekly Questions

### What is Quarto?

Quarto is a publishing system that mixes writing and code in one document.

- You write in Markdown.

- You embed R code in chunks.

- Quarto runs the code and weaves the output.

### How is Quarto different than a from-scratch HTML website?

- Quarto helps us write HTML website without writing HTML code!

- As a result, you don't have as much control over your website.

## `dplyr`

`dplyr` uses different verbs to work with tibbles.

- `filter()` picks rows

- `select()` picks columns

- `mutate()` adds a new computed column:

```{r}
#| eval: false
new <- mutate(counties, income2)
```

- `arrange()` changes the order of rows

- `group_by()` + `summarize()` many rows become few

For example:

| County     | State | Income  |
|------------|------:|---------|
| Orange     |    NY | 70,000  |
| Nassau     |    NY | 80,000  |
| Orange     |    CA | 90,000  |
| Sacramento |    CA | 100,000 |

Becomes:

| State | Income |
|-------|--------|
| NY    | 70,000 |
| CA    | 80,000 |

### Pipe operator `|>`

We can think of `|>` as saying "and then..."

To use the pipe operator, we use ⌘ + Shift + M.

```{r}
#| eval: false

new_data <- counties |> 
  filter(moe_pct > 10) |> 
  arrange(desc(moe_pct)) |> 
  select(NAME, moe_pct)
```

## The Census

**Decennial Census**: Mandated by the Constitution, conducted every 10 years with 10 questions.

**American Community Survey (ACS)**: Created in 2005, more interesting questions! Approximately 3% of households are surveyed about income, rent, home value, commute time, ethnicity, etc. From that sample, we come up with an estimate for the population.

- **1-year ACS**: Almost never used, only for places \> 65,000 people. Most current, comes out every year.

- **5-year ACS**: Block Group, Census Tract. If you get ACS data for the year 2016, it is a 5 year average. These are estimates +- a margin of error. It is **NOT** just an estimate for 2016, but for the interval of 2011-2016.

### Census Code

#### B19013_001E or B19013_001M

- B: Base table

- 19013: Table \#

- 001: Line Number

- E: Estimate

- M: Margin of Error

The command `load_variables()` shows us the name of our variables. All of this exists in a library called `tidycensus()`.

### Census Geography

- State -\> County -\> Tract ("Neighborhood") -\> Block Groups (HORRIBLE margins of error!) -\> Block (Decennial Census only)

- About 5,000-8,000 people

- Boundaries of census tracts change over time! To mitigate this, there are longitudinal tract databases

### GEOID

A GEOID is a unique identifier for a census tract.

- First two numbers is the state

- Next three numbers is the county

- Last six digits are the census tract

A GEOID is stored as string. This is important, as there are states such as Alabama with a `01` state identifier.

### Using `tidycensus()`

```{r}
#| eval: false
data <- get_acs(geography: 'county',
                variables = c('B1013_001',
                              'B01003_001'),
                state = 'PA',
                year = 2023,
                survey = 'acs5',
                output = 'wide') #To get our data in wide format 
```

When we run this command, we receive it in **tidy** format, with multiple rows for each county (each row shows a different variable value –\> "variable" itself is a column). We want to turn it into **wide** format, with different variables in each column.
