---
title: "Assignment 2: Who Do We Know the Least About?"
description: "Measuring uncertainty in American Community Survey data"
author: Michael Pincus
date: 09-04-2026
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
editor: 
  markdown: 
    wrap: 72
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

## The question

Every number the American Community Survey publishes is an estimate
drawn from a sample, not a count. That means every number comes with a
margin of error — a statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

**In this analysis, I look at total commuters across counties in
Connecticut, and ask which communities we measure most and least
reliably. As Connecticut features a large commuter region to New York in
the southwestern part of the state, I am curious about the general
patterns across Connecticut and where census estimates tend to be more
reliable.**

## Choosing what to measure

```{r}
my_state <- "CT"        # two-letter abbreviation, in quotes
my_variable <- "B08303_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```

**My data has 9 rows, matching the 9 census-used planning regions in the
state of Connecticut.**

------------------------------------------------------------------------

## How big is the margin of error?

A margin of error of 3,000 means something very different for an
estimate of 20,000 than for an estimate of 200,000. To compare across
places, we need it as a percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```

**The Northeastern Connecticut Planning Region has the largest `moe_pct`
relative to its estimate, measuring at about 2.6%.**

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 1  ~ "High confidence",
    moe_pct < 2 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

As the percent margin of error for all 9 planning regions of Connecticut
fell well into the original \<5% cutoff for `High confidence`, the
categories were adjusted so that `High confidence` represents a margin
of error of \<1%, `Moderate` represents a margin of error of \<2%, and
`Low confidence` represents a margin of error of \>2%.

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the
same row. That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```

```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 1  ~ "High confidence",
    moe_pct < 2 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability
category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  ) %>%
  arrange(desc(avg_population))
```

**The `avg_population` of high-confidence planning regions are much
higher than the `avg_population` of low-confidence planning regions, the
former being about 969,029 and the latter being about 104,338.**

## What this means for policy

Through this exercise, we learned that we know the least about the
Northeastern Connecticut and the Northeast Hills planning regions. As
both of these planning regions feature the lowest populations in the
state, they feature the lowest sample sizes and largest percent margin
of error (\>2%) in the American Community Survey data set.

It is important to understand the variability of reliability within the
ACS data set when making policy decisions, as certain planning regions
with lower sample sizes may be underrepresented. For example,
communities in the northeastern and northwestern corners of the state
may appear to have fewer commuters than in the Hartford region, though
in reality it may be an artifact of sample size. Poor data literacy in
the policy making process could result in a reduction in bus service
within these regions, despite them in reality having a larger need for
public transportation than anticipated.

## Plotting Data

```{r}
library(sf)
planning_regions <- read_sf('assignments/data/CT_Planning_Regions.geojson')

planning_regions <- left_join(planning_regions, county_wide, by=c('PlanningRegionFIPS_GEOID'='GEOID'))

planning_regions<- planning_regions %>%
  arrange(match(reliability, c("High confidence", "Moderate", "Low confidence")))
```

```{r}
library(patchwork)
reliability_map <- ggplot(planning_regions) +
  geom_sf(aes(fill=reliability), color='white') +
  scale_fill_manual(values = c("High confidence" = "#b71c1c", 
                               "Moderate" = "#f4977a", 
                               "Low confidence" = "#fce4de"),
                    breaks = c("High confidence", "Moderate", "Low confidence")) + 
  labs(title='Reliability of Census Estimates', fill=NULL) +
  theme_void()

pop_map <- ggplot(planning_regions) +
  geom_sf(aes(fill = popE), color = 'white') +
  scale_fill_distiller(palette = "Reds", direction = 1) +
  labs(title='Population', fill=NULL) + 
  theme_void()

reliability_map + pop_map
```
